---
title: "Version 6"
author: "J. Ignacio Guzmán Serrano (685935jg)"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Libraries

```{r}
library(sampleSelection)
library(mvtnorm)
library(caret)
library(ggplot2)
library(rgl)
library(Metrics)
library(truncnorm)
library(knitr)
library(sandwich)
library(kableExtra)
library(MASS)
library(nortest)
library(moments)
library(tseries)
```

### Read in advance

Check the Outline. There are 3 parts:

1.  Creation of data set 'data': I used the format code by Toomet & Henningsen (2008). The first chunck (1.1) creates a data set when variables are not correlated and in the second chunck (1.2) when you want to correlate them. I could have used just the second chunck for everything and set correlation to 0 when not wanting variables to be correlated but I took the first option

    No matter the option, in this section the different combinations for predictions are created: the scenarios, censoring, sample size, correlation of errors and correlation of variables.

    (The one by default is Scenario 1, sample size = 800, correlation errors = 0, degree of censoring = 30%)

2.  Running iterations: 100 iterations where models are trained and validated. WARNING: change % partition depending on sample size (0.625 for sample size = 800; and 0.869 when sample size = 2300) and change Heckman models code when trained depending on the Scenario to be tested.

3.  Results: all in one chunck that will generated:

    -   A table with metrics for predictions made on 'test_ws'.

    -   A second table with metrics for predictions made on 'test'.

    -   A third table with the coefficients from models.

    -   Two plots. A boxplot of all the MSE from predictions made on 'test_ws'. And a plot with MSE from predictions on 'test_ws' from each iteration.

# 1. Creation of data set 'data'

#### 1.1 When variables not correlated

```{r}
set.seed(22)

# --------------------------------------------------------------------

v1 <- runif(800) # Random values from 0 to 1
v2 <- runif(800) # Random values from 0 to 1
errors <- rmvnorm(800, c(0, 0), matrix(c(1, 0,0, 1), 2, 2))
                 # Errors correlated at -0.7

# --------------------------------------------------------------------

# join previous variables to a data frame.
data <- data.frame(var1 = v1, var2 = v2, errors = errors)
# add to that data frame the following variables:
data$selection <- data$var2 + data$errors.2 > 0
data$latent.truth <- data$var1 + data$errors.1
data$bias.outcome <- data$latent.truth * (data$selection > 0)

rm( v1, v2, errors)

# Check censoring degree
(nrow(data[data$bias.outcome == 0, ])/nrow(data))
```

#### 1.2 When variables correlated

```{r}
set.seed(22)

# --------------------------------------------------------------------


correlated_normals <- mvrnorm(800, c(0, 0), matrix(c(1, -0.5, -0.5, 1), 2, 2))

# Transform normal variables to uniform variables
v1 <- pnorm(correlated_normals[,1])  # Transform to uniform
v2 <- pnorm(correlated_normals[,2])  # Transform to uniform
errors <- rmvnorm(800, c(0, 0), matrix(c(1, -0.85,-0.85, 1), 2, 2))

# --------------------------------------------------------------------
# join previous variables to a data frame.
data <- data.frame(var1 = v1, var2 = v2, errors = errors)
# add to that data frame the following variables:
data$selection <- data$var2 +  data$errors.2 > 0
data$latent.truth <- data$var1 + data$errors.1
data$bias.outcome <- data$latent.truth * (data$selection > 0)

rm( v1, v2, errors)
# Check censoring degree
(nrow(data[data$bias.outcome == 0, ])/nrow(data))
```

# 2. Iterations

Predictions are done 2 times, one with 'test' set and another one with 'test_ws' set. - The ones done with 'test' are recorded with letter **A** to differentiate.

```{r}


# Initialize empty matrices to store results -----------------------------------
num_models <- 4
num_metrics <- 5
num_tests <- 4
mae <- matrix(NA, nrow = 100, ncol = num_models)
mse <- matrix(NA, nrow = 100, ncol = num_models)
rmse <- matrix(NA, nrow = 100, ncol = num_models)
var_resid <- matrix(NA, nrow = 100, ncol = num_models)
bias_resid <- matrix(NA, nrow = 100, ncol = num_models)

coefficients_ols <- data.frame(Iteration = numeric(), Intercept = numeric(), Var2 = numeric(), SE_Intercept = numeric(), SE_Var2 = numeric())
coefficients_ml <- data.frame(Iteration = numeric(), Intercept = numeric(), Var2 = numeric(), SE_Intercept = numeric(), SE_Var2 = numeric())
coefficients_2step <- data.frame(Iteration = numeric(), Intercept = numeric(), Var2 = numeric(), SE_Intercept = numeric(), SE_Var2 = numeric())
coefficients_true <- data.frame(Iteration = numeric(), Intercept = numeric(), Var2 = numeric(), SE_Intercept = numeric(), SE_Var2 = numeric())

mse_true <- data.frame(Iteration = numeric(), MSE = numeric())
mse_ols <- data.frame(Iteration = numeric(), MSE = numeric())
mse_ml <- data.frame(Iteration = numeric(), MSE = numeric())
mse_2step <- data.frame(Iteration = numeric(), MSE = numeric())

A.mae <- matrix(NA, nrow = 100, ncol = num_models)
A.mse <- matrix(NA, nrow = 100, ncol = num_models)
A.rmse <- matrix(NA, nrow = 100, ncol = num_models)
A.var_resid <- matrix(NA, nrow = 100, ncol = num_models)
A.bias_resid <- matrix(NA, nrow = 100, ncol = num_models)


A.mse_true <- data.frame(Iteration = numeric(), MSE = numeric())
A.mse_ols <- data.frame(Iteration = numeric(), MSE = numeric())
A.mse_ml <- data.frame(Iteration = numeric(), MSE = numeric())
A.mse_2step <- data.frame(Iteration = numeric(), MSE = numeric())


jb_p <- matrix(NA, nrow = 100, ncol = num_models)
jb_s <- matrix(NA, nrow = 100, ncol = num_models)
rescaled_moments_p <- matrix(NA, nrow = 100, ncol = num_models)
censoring <- matrix(NA, nrow = 100)




# Loop for 100 iterations -----------------------------------------------------

for (i in 1:100) {
  set.seed(i)  # Set a different seed for each iteration
  
  # Create training and test set
  train_indices <- createDataPartition(data$var1, p = 0.625, list = FALSE)
  train <- data[train_indices, ]
  test <- data[-train_indices, ]
  
  # Subset data for Heckman model
  train_ws <- train[train$selection, ]
  test_ws <- test[test$selection, ]
  
  
  censoring[i,] <- 1-(nrow(train[train$selection, ])/nrow(train)) # calculate the degree of censoring
  
  # Fit models
  model.true <- lm(latent.truth ~ var1, data = train)
  model.ols <- lm(latent.truth ~ var1, data = train_ws)
  model.ml <- selection(selection ~ var2 , bias.outcome ~ var1, 
                        method = "ml", data = train)
  model.2step <- selection(selection ~ var2, bias.outcome ~ var1, 
                           method = "2step", data = train)
  
   # Store coefficients and standard errors
  coef_true <- coef(summary(model.true))
  coef_ols <- coef(summary(model.ols))
  coef_ml <- coef(summary(model.ml))
  coef_2step <- coef(summary(model.2step))
 
  
  

# MAKING PREDICTIONS WITH 'test' set (all individuals, observed and unobserved) ----------------------------------------- 
  
  # Make predictions
  A.p.true <- predict(model.true, newdata = test)
  A.p.ols <- predict(model.ols, newdata = test)
  A.p.ml <- predict(model.ml, newdata = test)
  A.p.2step <- predict(model.2step, newdata = test)
  
  # Compute metrics
  A.residuals_true <- test$latent.truth - A.p.true
  A.residuals_ols <- test$latent.truth - A.p.ols
  A.residuals_ml <- test$latent.truth - A.p.ml
  A.residuals_2step <- test$latent.truth - A.p.2step
  
  A.mae[i, ] <- c(mean(abs(A.residuals_ols)), mean(abs(A.residuals_ml)), mean(abs(A.residuals_2step)), mean(abs(A.residuals_true)))
  A.mse[i, ] <- c(mean(A.residuals_ols^2), mean(A.residuals_ml^2), mean(A.residuals_2step^2), mean(A.residuals_true^2))
  A.rmse[i, ] <- c(sqrt(mean(A.residuals_ols^2)), sqrt(mean(A.residuals_ml^2)), sqrt(mean(A.residuals_2step^2)), sqrt(mean(A.residuals_true^2)))
  A.var_resid[i, ] <- c(var(A.residuals_ols), var(A.residuals_ml), var(A.residuals_2step), var(A.residuals_true))
  A.bias_resid[i, ] <- c((mean(A.p.ols) - mean(test$latent.truth))^2, 
                       (mean(A.p.ml) - mean(test$latent.truth))^2,
                       (mean(A.p.2step) - mean(test$latent.truth))^2, 
                       (mean(A.p.true) - mean(test$latent.truth))^2)
  
  # Append MSE for each model to respective data frames
  A.mse_true <- rbind(A.mse_true, data.frame(Iteration = i, MSE = mean(A.residuals_true^2)))
  A.mse_ols <- rbind(A.mse_ols, data.frame(Iteration = i, MSE = mean(A.residuals_ols^2)))
  A.mse_ml <- rbind(A.mse_ml, data.frame(Iteration = i, MSE = mean(A.residuals_ml^2)))
  A.mse_2step <- rbind(A.mse_2step, data.frame(Iteration = i, MSE = mean(A.residuals_2step^2)))
  
  
  
  
  
# MAKING PREDICTIONS WITH 'TEST WITH  OBSERVED'test_ws' (only selected individuals) ----------------------------------------- 
  # Make predictions
  p.true <- predict(model.true, newdata = test_ws)
  p.ols <- predict(model.ols, newdata = test_ws)
  p.ml <- predict(model.ml, newdata = test_ws)
  p.2step <- predict(model.2step, newdata = test_ws)
  
  # Compute metrics
  residuals_true <- test_ws$latent.truth - p.true
  residuals_ols <- test_ws$latent.truth - p.ols
  residuals_ml <- test_ws$latent.truth - p.ml
  residuals_2step <- test_ws$latent.truth - p.2step
  
  mae[i, ] <- c(mean(abs(residuals_ols)), mean(abs(residuals_ml)), mean(abs(residuals_2step)), mean(abs(residuals_true)))
  mse[i, ] <- c(mean(residuals_ols^2), mean(residuals_ml^2), mean(residuals_2step^2), mean(residuals_true^2))
  rmse[i, ] <- c(sqrt(mean(residuals_ols^2)), sqrt(mean(residuals_ml^2)), sqrt(mean(residuals_2step^2)), sqrt(mean(residuals_true^2)))
  var_resid[i, ] <- c(var(residuals_ols), var(residuals_ml), var(residuals_2step), var(residuals_true))
  bias_resid[i, ] <- c((mean(p.ols) - mean(test_ws$latent.truth))^2, 
                       (mean(p.ml) - mean(test_ws$latent.truth))^2,
                       (mean(p.2step) - mean(test_ws$latent.truth))^2, 
                       (mean(p.true) - mean(test_ws$latent.truth))^2)
  
  # Append MSE for each model to respective data frames
  mse_true <- rbind(mse_true, data.frame(Iteration = i, MSE = mean(residuals_true^2)))
  mse_ols <- rbind(mse_ols, data.frame(Iteration = i, MSE = mean(residuals_ols^2)))
  mse_ml <- rbind(mse_ml, data.frame(Iteration = i, MSE = mean(residuals_ml^2)))
  mse_2step <- rbind(mse_2step, data.frame(Iteration = i, MSE = mean(residuals_2step^2)))

  
  

#---------------------------------------------------------------------------
  # Append coefficients for each model to respective data frames
  coefficients_ols <- rbind(coefficients_ols, data.frame(Iteration = i, Intercept = coef_ols[1,1], Var2 = coef_ols[2,1], SE_Intercept = coef_ols[1,2], SE_Var2 = coef_ols[2,2]))
  coefficients_ml <- rbind(coefficients_ml, data.frame(Iteration = i, Intercept = coef_ml[1,1], Var2 = coef_ml[2,1], SE_Intercept = coef_ml[1,2], SE_Var2 = coef_ml[2,2]))
  coefficients_2step <- rbind(coefficients_2step, data.frame(Iteration = i, Intercept = coef_2step[1,1], Var2 = coef_2step[2,1], SE_Intercept = coef_2step[1,2], SE_Var2 = coef_2step[2,2]))
  coefficients_true <- rbind(coefficients_true, data.frame(Iteration = i, Intercept = coef_true[1,1], Var2 = coef_true[2,1], SE_Intercept = coef_true[1,2], SE_Var2 = coef_true[2,2]))

  # Normality test (Jarque-Bera)
  jb_p[i, ] <- c(jarque.bera.test(residuals_ols)$p.value, 
                 jarque.bera.test(residuals_ml)$p.value, 
                 jarque.bera.test(residuals_2step)$p.value,
                 jarque.bera.test(residuals_true)$p.value)
  jb_s[i, ] <- c(jarque.bera.test(residuals_ols)$statistic, 
                 jarque.bera.test(residuals_ml)$statistic, 
                 jarque.bera.test(residuals_2step)$statistic,
                 jarque.bera.test(residuals_true)$statistic)
  
  # Rescaled moments normality test
  skewness_values <- c(skewness(residuals_ols), skewness(residuals_ml), skewness(residuals_2step), skewness(residuals_true))
  kurtosis_values <- c(kurtosis(residuals_ols), kurtosis(residuals_ml), kurtosis(residuals_2step), kurtosis(residuals_true))
  
  n_values <- c(length(residuals_ols), length(residuals_ml), length(residuals_2step), length(residuals_true))
  k_values <- c(length(model.ols$coefficients) - 1, length(model.ml$coefficients) - 1, length(model.2step$coefficients) - 1, length(model.true$coefficients) - 1)
  
  c_values <- n_values / (n_values - k_values)
  rescaled_skewness <- sqrt(c_values) * skewness_values
  rescaled_kurtosis <- c_values * (kurtosis_values - 3) + 3
  
  JB <- (n_values / 6) * (rescaled_skewness^2 + (rescaled_kurtosis - 3)^2 / 4)
  rescaled_moments_p[i, ] <- pchisq(JB, df = 2, lower.tail = FALSE)

  
  
  
  }
```

# 3. Metrics

```{r}
print(average.censoring <- mean(censoring))
```

```{r}
# Calculate average of each metric
avg_mae <- colMeans(mae)
avg_mse <- colMeans(mse)
avg_rmse <- colMeans(rmse)
avg_var_resid <- colMeans(var_resid)
avg_bias_resid <- colMeans(bias_resid)
avg_jb_p <- colMeans(jb_p)
avg_jb_s <- colMeans(jb_s)
avg_rescaled_moments_p <- colMeans(rescaled_moments_p)

# Create a table to display the results
results <- data.frame(
  Model = c("OLS", "ML", "2Step","True"),
  MAE = avg_mae,
  RMSE = avg_rmse,
  MSE = avg_mse,
  Var.residuals = avg_var_resid,
  Sqr_bias.residuals = avg_bias_resid,
  Jarque_Bera.statistic = avg_jb_s,
  Jarque_Bera.pvalue = avg_jb_p,
  Rescale_moments.pvalue = avg_rescaled_moments_p
  
)

print(results)

# Calculate average of each metric
A.avg_mse <- colMeans(A.mse)
A.avg_var_resid <- colMeans(A.var_resid)
A.avg_bias_resid <- colMeans(A.bias_resid)

# Create a table to display the results
A.results <- data.frame(
  Model = c("OLS", "ML", "2Step","True"),
  MSE = A.avg_mse,
  Var.residuals = A.avg_var_resid,
  Sqr_bias.residuals = A.avg_bias_resid
  
)

print(A.results)

# Compute the average coefficients for each model
average_coefficients_ols <- aggregate(. ~ Iteration, coefficients_ols, mean)
average_coefficients_ml <- aggregate(. ~ Iteration, coefficients_ml, mean)
average_coefficients_2step <- aggregate(. ~ Iteration, coefficients_2step, mean)
average_coefficients_true <- aggregate(. ~ Iteration, coefficients_true, mean)

# Display the average coefficients and standard errors
average_coefficients <- data.frame(
  Model = c("OLS", "ML", "Two-Step","true"),
  Intercept = c(mean(average_coefficients_ols$Intercept), mean(average_coefficients_ml$Intercept), mean(average_coefficients_2step$Intercept), mean(average_coefficients_true$Intercept)),
  SE_Intercept = c(mean(average_coefficients_ols$SE_Intercept), mean(average_coefficients_ml$SE_Intercept), mean(average_coefficients_2step$SE_Intercept), mean(average_coefficients_true$SE_Intercept)),
  Var1 = c(mean(average_coefficients_ols$Var2), mean(average_coefficients_ml$Var2), mean(average_coefficients_2step$Var2), mean(average_coefficients_true$Var2)),
  SE_Var1 = c(mean(average_coefficients_ols$SE_Var2), mean(average_coefficients_ml$SE_Var2), mean(average_coefficients_2step$SE_Var2), mean(average_coefficients_true$SE_Var2))
)

kable(average_coefficients, format = "html") %>%
  kable_styling(full_width = FALSE)

# Plot MSE for each iteration for each model
plot(mse_ols$Iteration, mse_ols$MSE, type = "l", col = "blue",
     xlab = "Iteration", ylab = "Mean Squared Error", main = "MSE for Each Model",
     ylim = range(c(mse_ols$MSE, mse_ml$MSE, mse_2step$MSE)))
lines(mse_ml$Iteration, mse_ml$MSE, type = "l", col = "red")
lines(mse_2step$Iteration, mse_2step$MSE, type = "l", col = "green")

# Add legend
legend("topright", legend = c("OLS", "ML", "Two-Step"), col = c("blue", "red", "green"), lty = 1)



# Create a matrix of MSE values for each model
mse_matrix <- cbind(mse[, 1], mse[, 2], mse[, 3])

# Convert the matrix to a data frame
mse_data <- as.data.frame(mse_matrix)
colnames(mse_data) <- c("OLS", "ML", "2Step")

# Create the boxplot
boxplot(mse_data, col = c("blue", "red", "green"), 
        main = "MSE for Each Model", ylab = "MSE", xlab = "Model")

```
